---
title: "Assignment 2"
author: "anonymous"
pdf:
  geometry: default
  number-sections: yes
  code-annotations: none
output: pdf_document
---




Loading the library and the data.

```{r}
library(aaltobda)
data("algae")
# The data are now stored in the variable `algae`.
# These are the values for the prior required in the assignment
prior_alpha = 2
prior_beta = 10
```

The below data is **only for the tests**, you need to change to the full data `algae` when reporting your results.

```{r}
algae_test <- c(0, 1, 1, 0, 0, 0)
```

## (a)

Write the likelihood, the prior and the posterior here!
```{r}
y = 44
n = 274
p = seq(0,1,length=100)
```

Likelihood 
\begin{align*}
$$
p(y |\pi) = \binom{n}{y} \pi^x (1-\pi)^{n-y}
$$
\end{align*}

Prior 

\begin{align*}

$$
P(\theta) = \frac{1}{\text{Beta}(\alpha, \beta)} \theta^ {\alpha - 1} (1 - \theta)^{\beta - 1}
$$
\end{align*}

Posterior
\begin{align*}
$$
\pi | y \sim \text{Beta}(y + \alpha, n - y + \beta)
$$
\end{align*}

Since there has been recorded 44 instances of algae in the dataset, we set Y = 44 and since the number of observations is 274, we get the following parameters for the posterior

\begin{align*}
\pi | y \sim \text{Beta}(46, 240)
\end{align*}


## (b)

Keep the below name and format for the functions to work with `markmyassignment`:


```{r}
# Useful function: qbeta()

beta_point_est <- function(prior_alpha, prior_beta, data) {
    y = sum(data)
    n = length(data)
    posterior_alpha = prior_alpha + y
    posterior_beta = prior_beta + (n-y)
    posterior_mean = posterior_alpha/(posterior_beta+posterior_alpha)
    return(posterior_mean)
}
beta_interval <- function(prior_alpha, prior_beta, data, prob=0.9) {
    y = sum(data)
    n = length(data)
    posterior_alpha = prior_alpha + y
    posterior_beta = prior_beta + (n-y)
    upper = quantile(qbeta(p, posterior_alpha, posterior_beta), prob = 0.95)
    lower = quantile(qbeta(p, posterior_alpha, posterior_beta), prob = 0.05)
    
    
    return(c(lower,upper))
}

print(beta_point_est(prior_alpha, prior_beta, algae))
print(beta_interval(prior_alpha, prior_beta, algae, prob = 0.9))
```

The posterior mean estimate is roughly 0.16
The posterior interval is from 0.126 to 0.197
## (c)

Keep the below name and format for the function to work with `markmyassignment`:

```{r}
# Useful function: pbeta()

beta_low <- function(prior_alpha, prior_beta, data, pi_0=0.2) {
    y = sum(data)
    n = length(data)
    posterior_alpha = prior_alpha + y
    posterior_beta = prior_beta + (n-y)
    return(pbeta(pi_0,posterior_alpha, posterior_beta))
}

print(beta_low(prior_alpha, prior_beta, algae, pi_0 = 0.2))
```
The  probability that the proportion of monitoring sites with detectable algae levels is smaller than 0.2 equals to roughly 0.96


## (d)

Write your answer here!

- The presence of algae in a lake should be independent from the presence of algae in other lakes
- We should assume that the probability of algae being present is equal across all lakes
- We should have enough data to make conclusions based on the data, for example a dataset of 3 observations will not take us far



## (e)

Plot the PDFs here. Explain shortly what you do.

Let's first create a few new prior values for the prior beta function.
```{r}
# Useful function: dbeta()
y = 44
n = 274

#The prior values used earlier in the assignment
prior_alpha = 2
prior_beta = 10

posterior_alpha = prior_alpha + y
posterior_beta = prior_beta + (n-y)

#Let's create 3 separate prior values for comparison
prior_alpha_1 = 6
prior_beta_1 = 24

posterior_alpha_1 = prior_alpha_1+ y
posterior_beta_1 = prior_beta_1 + (n-y)

prior_alpha_2 = 18
prior_beta_2 = 100

posterior_alpha_2 = prior_alpha_2 + y
posterior_beta_2 = prior_beta_2 + (n-y)

prior_alpha_3 = 30
prior_beta_3 = 150

posterior_alpha_3 = prior_alpha_3 + y
posterior_beta_3 = prior_beta_3 + (n-y)





#The posteriors plotted with four different prior parameters
plot(p, dbeta(p,posterior_alpha, posterior_beta), type = "l", col = "red", ylab = "Density", xlab = expression(pi), ylim = c(0,25))
lines(p, dbeta(p,posterior_alpha_1, posterior_beta_1), type = "l", col = "blue")
lines(p, dbeta(p,posterior_alpha_2, posterior_beta_2), type = "l", col = "purple")
lines(p, dbeta(p,posterior_alpha_3, posterior_beta_3), type = "l", col = "green")
legend("topright", legend = c("Posterior with the prior Beta(2,10)","Posterior with the prior Beta(6,24)","Posterior with the prior Beta(18,100)","Posterior with the prior Beta(30,150)"), col = c("red", "blue", "purple", "green"), lty = c(1))
```

As we can see from the plotted figure, the posterior seems to be very consistent regardless of the parameters given to the prior. 


## Use of AI

I used AI in the report to find out errors in my Latex syntax, also there was an issue with the pdf formatting issues, that were solved using AI. 
