---
title: "Assignment 3"
author: "anonymous"
format:
  html:
    toc: yes
    code-tools: yes
    code-line-numbers: yes
    number-sections: yes
    mainfont: Georgia, serif
    page-layout: article
  pdf:
    geometry: left=1cm,top=1cm,bottom=1cm,right=7cm
    number-sections: yes
    code-annotations: none
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
editor: source
---

# General information

::: {.content-hidden when-format="pdf"}
::: {.callout-warning collapse="false"}
## Setup

*This block will only be visible in your HTML output, but will be hidden
when rendering to PDF with quarto for the submission.* **Make sure that
this does not get displayed in the PDF!**

This is the template for [assignment 3](assignment3.html). You can
download the qmd-files
([full](https://avehtari.github.io/BDA_course_Aalto/assignments/template3.qmd),
[simple](https://avehtari.github.io/BDA_course_Aalto/assignments/simple_template3.qmd))
or copy the code from this rendered document after clicking on
`</> Code` in the top right corner.

**Please replace the instructions in this template by your own text,
explaining what you are doing in each exercise.**

The following will set-up
[`markmyassignment`](https://github.com/MansMeg/markmyassignment) to
check your functions at the end of the notebook:

The following installs and loads the `aaltobda` package:

```{r}
if(!require(aaltobda)){
    install.packages("remotes")
    remotes::install_github("avehtari/BDA_course_Aalto", subdir = "rpackage", upgrade="never")
    library(aaltobda)
}
```

The following installs and loads the [`latex2exp`
package](https://github.com/stefano-meschiari/latex2exp), which allows
us to use LaTeX in plots:

```{r}
if(!require(latex2exp)){
    install.packages("latex2exp")
    library(latex2exp)
}
```
:::
:::

::: {.content-hidden when-format="pdf"}
::: {.callout-tip collapse="false"}
## Showcase: Setting up advanced packages (`posterior` and `ggdist`)

*This block will only be visible in your HTML output, but will be hidden
when rendering to PDF with quarto for the submission.* **Make sure that
this does not get displayed in the PDF!**

*This block showcases advanced tools, which you will be allowed and
expected to use after this assignment.* **For now, you should solve the
assignment without the tools showcased herein.**

The following installs and loads the [`posterior`
package](https://mc-stan.org/posterior/index.html), which allows us to
use its [`rvar` Random Variable
Datatype](https://mc-stan.org/posterior/articles/rvar.html):

```{r}
if(!require(posterior)){
    install.packages("posterior")
    library(posterior)
}
```

The following installs and loads the [`ggdist`
package](https://mjskay.github.io/ggdist/) for advanced plotting
functions:

```{r}
if(!require(ggplot2)){
    install.packages("ggplot2")
    library(ggplot2)
}
ggplot2::theme_set(theme_minimal(base_size = 14))
if(!require(ggdist)){
    install.packages("ggdist")
    library(ggdist)
}
```

*This block showcases advanced tools, which you will be allowed and
expected to use after this assignment.* **For now, you should solve the
assignment without the tools showcased herein.**
:::
:::

# Inference for normal mean and deviation (3 points)

Loading the library and the data.

```{r}
data("windshieldy1")
# The data are now stored in the variable `windshieldy1`.
# The below displays the data:
windshieldy1
```

The below data is **only for the tests**, you need to change to the full
data `windshieldy1` when reporting your results.

```{r}
windshieldy_test <- c(13.357, 14.928, 14.896, 14.820)
```

## (a)

Write your answers here!

Likelihood:


$$
L(\mu, \sigma^2|y) = \prod_{i=1}^{n} \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{(y_i - \mu)^2}{2\sigma^2}\right) 
$$

Prior


$$
p(µ, \sigma^2) \propto (\sigma^2)^{-1}
$$

Posterior

$$
p(\mu, \sigma^2 | y) = \sigma^{-n-2} \exp\left(-\frac{1}{2\sigma^2} \left[(n-1)s^2 + n(y - \mu)^2\right]\right)
$$
## (b)

Write your answers and code here!

**Keep the below name and format for the functions to work with
`markmyassignment`:**

```{r}
# Useful functions: mean(), length(), sqrt(), sum()
# and qtnew(), dtnew() (from aaltobda)

mu_point_est <- function(data) {
    
    return(mean(data))
    
}
mu_interval <- function(data, prob = 0.95) {
    n = length(data)
    avg  = mean(data)
    df = n-1
    s = sd(data)

    alpha = 1-prob
    t = qtnew(1- alpha/2, df)  
    se <- s / sqrt(n)
    margin = se * t
    interval_lower = avg-margin
    interval_higher = avg + margin
    return(c(interval_lower, interval_higher))
    
}

print(mu_point_est(windshieldy1))
print(mu_interval(windshieldy1, prob = 0.95))


```

The value of the point estimate is 14.61 and the value for the interval
limits are [13.48, 15.744]

You can plot the density as below if you implement `mu_pdf` to compute
the PDF of the posterior $p(\mu|y)$ of the average hardness ${\mu}$.

```{r}
#| label: fig-2b-density
#| fig-cap: PDF of the posterior $p(\mu|y)$ of the average hardness $\mu$
mu_pdf <- function(data, x){
    # Compute necessary parameters here.
    # These are the correct parameters for `windshieldy_test` 
    # with the provided uninformative prior.
    n = length(data)
    df = n-1
    location = mean(data)
    scale = sqrt(sd(data)^2/length(data))
    # Use the computed parameters as below to compute the PDF:
    print(scale)
    return(dtnew(x, df, location, scale))
}

conf_interval = mu_interval(windshieldy1, .999)

lower = conf_interval[1]
upper= conf_interval[2]
x = seq(lower, upper, length.out=1000)
plot(
    x, mu_pdf(windshieldy1, x), type="l", 
    xlab=TeX(r'(average hardness $\mu$)'), 
    ylab=TeX(r'(PDF of the posterior $p(\mu|y)$)')
)
```

Based on these results we can conclude that the population mean $\mu$
will fall in between this interval with a 95% confidence.

## (c)

Write your answers and code here!

**Keep the below name and format for the functions to work with
`markmyassignment`:**

```{r}
# Useful functions: mean(), length(), sqrt(), sum()
# and qtnew(), dtnew() (from aaltobda)

mu_pred_point_est <- function(data) {
    # Do computation here, and return as below.
    # This is the correct return value for the test data provided above.
    return(mean(data))
    #14.5
    
}
mu_pred_interval <- function(data, prob = 0.95) {
  
  probability = prob
  df = length(data)-1
  avg = mean(data)
  s = sd(data)
  alpha = 1-probability
  t_value = qt(1-alpha/2, df)
  
 
  lower = avg-t_value * s * sqrt(1+1/length(data))
  upper = avg+t_value * s * sqrt(1+1/length(data))
  
  




    return(c(lower, upper))
}

print(mu_pred_point_est(windshieldy1))
print(mu_pred_interval(windshieldy1))
```

The predictions for the point estimate and confidence interval are the
following: 14.61 and \[11.03,18.2\]

You can plot the density as below if you implement `mu_pred_pdf` to
compute the PDF of the posterior predictive $p(\tilde{y}|y)$ of a new
hardness observation $\tilde{y}$.

```{r}
#| label: fig-2c-density
#| fig-cap: PDF of the posterior predictive $p(\tilde{y}|y)$ of a new hardness observation $\tilde{y}$
mu_pred_pdf <- function(data, x){
    # Compute necessary parameters here.
    # These are the correct parameters for `windshieldy_test` 
    # with the provided uninformative prior.
    df = length(data)-1
    location = mean(data)
    scale = sd(data)*(sqrt(1+1/length(data)))
    print(scale)
    #0.8536316
    # Use the computed parameters as below to compute the PDF:
     
    dtnew(x, df, location, scale)
}

x_interval = mu_pred_interval(windshieldy1, .999)
lower_x = x_interval[1]
upper_x = x_interval[2]
x = seq(5, 25, length.out=1000)
plot(
    x, mu_pred_pdf(windshieldy1, x), type="l", 
    xlab=TeX(r'(new hardness observation $\tilde{y}$)'), 
    ylab=TeX(r'(PDF of the posterior predictive $p(\tilde{y}|y)$)')
)
```

# Inference for the difference between proportions (3 points)

## (a)

Write your answers here! Likelihood:

```{=tex}
\begin{align*}
p(y|\theta) &\propto \prod_{j=1}^{k} \theta^{y_j}
\end{align*}
```
Prior
We're setting a uniform prior of $\alpha_1 = 1$ for all $j$.
```{=tex}
\begin{align*}
p(\theta | \alpha) &\propto \prod_{j=1}^{k} \theta^{\alpha_j - 1}
\end{align*}
```
Posterior

Written in the course book : "The resulting posterior distribution for
the ${\theta_j}$ 's is Dirichlet with parameters ${\alpha_j}$ + ${y_j}$

With a noninformative uniform prior distribution on ${\theta, \alpha1 = \alpha2 = 1}$, the
posterior distributions for ${(\theta1, \theta2)}$is Dirichlet(40, 675) and Dirichlet(23,681)

## (b)

Write your answers and code here!

The below data is **only for the tests**:

```{r}
set.seed(4711)
ndraws = 1000
p0 = rbeta(ndraws, 5, 95)
p1 = rbeta(ndraws, 10, 90)

```

**Keep the below name and format for the functions to work with
`markmyassignment`:**

```{r}

deaths_p0 = 39 
deaths_p1 = 22
total_p0 = 674
total_p1 = 680


prior_alpha = 1
prior_beta = 1


set.seed(4711)
ndraws = 1000
p0 = rbeta(ndraws, deaths_p0+prior_alpha, total_p0+prior_beta)
p1 = rbeta(ndraws, deaths_p1+prior_alpha, total_p1+prior_beta)


```

```{r}
# Useful function: mean(), quantile()

posterior_odds_ratio_point_est <- function(p0, p1) {
    # Do computation here, and return as below.
    # This is the correct return value for the test data provided above.
    or = (p1/(1-p1))/(p0/(1-p0))
    return (mean(or))
    
}
posterior_odds_ratio_interval <- function(p0, p1, prob = 0.95) {
    limit = (1-prob)/2
    or = (p1/(1-p1))/(p0/(1-p0))
    credible_interval <- quantile(or, c(limit, 1-limit))
    #c(0.6796942,7.3015964)
    
    return (credible_interval)
}
or = (p1/(1-p1))/(p0/(1-p0))
print(posterior_odds_ratio_point_est(p0,p1))
print(posterior_odds_ratio_interval(p0,p1, 0.95))
hist(or, main = " Odds Ratio Histogram", xlab = "Odds Ratio", col = "lightblue")

```

Based on these simulations of the control and treatment groups, we can
conclude that the odds ratio will fall inbetween values of 0.331 and
0.947 with a probability of 95%. The histogram displays that the values
concentrate heavily around an odds ratio of 0.5, like the calculated
point estimate of the odds ratio suggests.


## (c)

Assuming that in the rubric when referenced to computations in part a),
it was intended to reference to the computations in part b)

```{r}


prior_alpha_1 = 4
prior_beta_1 = 4
deaths_p0 = 39 
deaths_p1 = 22
total_p0 = 674
total_p1 = 680

prior_alpha_1 = 4
prior_beta_1 = 4
prior_alpha_2 = 8
prior_beta_2 = 8


set.seed(4711)
ndraws = 1000
p0_1 = rbeta(ndraws, deaths_p0+prior_alpha_1, total_p0+prior_beta_1)
p1_1 = rbeta(ndraws, deaths_p1+prior_alpha_1, total_p1+prior_beta_1)

p0_2 = rbeta(ndraws, deaths_p0+prior_alpha_2, total_p0+prior_beta_2)
p1_2 = rbeta(ndraws, deaths_p1+prior_alpha_2, total_p1+prior_beta_2)

or_1 = (p1_1/(1-p1_1))/(p0_1/(1-p0_1))
print(posterior_odds_ratio_point_est(p0_1,p1_1))
print(posterior_odds_ratio_interval(p0_1,p1_1, 0.95))
hist(or_1, main = " Odds Ratio Histogram", xlab = "Odds Ratio", col = "lightblue")

or_2 = (p1_2/(1-p1_2))/(p0_2/(1-p0_2))
print(posterior_odds_ratio_point_est(p0_2,p1_2))
print(posterior_odds_ratio_interval(p0_2,p1_2, 0.95))
hist(or_2, main = " Odds Ratio Histogram", xlab = "Odds Ratio", col = "lightblue")
```

Looking at the results from using different priors, we can see a clear
trend of the odds ratio moving towards 1. This means that the difference
between the survival rate between the control and treatment groups is
getting smaller. This is due to the probability of death increasing in
both groups.

# Inference for the difference between normal means (3 points)

Loading the library and the data.

```{r}
data("windshieldy2")
# The new data are now stored in the variable `windshieldy2`.
# The below displays the first few rows of the new data:
head(windshieldy2)
```

## (a)

Write your answers here!

Likelihood:

```{=tex}
\begin{align*}


L(\mu_1, \sigma_1, \mu_2, \sigma_2 | y_1, y_2) = \frac{1}{2\pi\sigma_1^2} \exp\left(-\frac{(y_1 - \mu_1)^2}{2\sigma_1^2}\right) \cdot \frac{1}{2\pi\sigma_2^2} \exp\left(-\frac{(y_2 - \mu_2)^2}{2\sigma_2^2}\right)

\end{align*}
```
Prior

```{=tex}
\begin{align*}

p(µ_i, \sigma²_i) \propto (\sigma^2_i)^{-1}

\end{align*}
```
Posterior

```{=tex}
\begin{align*}

p(\mu, \sigma^2 | y) \propto L(\mu_1, \sigma_1, \mu_2, \sigma_2 | y_1, y_2) * p(µ_1, \sigma^2_1) * p(µ_2, \sigma^2_2)

\end{align*}
```
## (b)

Write your answers and code here!

```{r}
# Useful functions: mean(), length(), sqrt(), sum(),
# rtnew() (from aaltobda), quantile() and hist().

y1  = windshieldy1
y2 = windshieldy2


  
set.seed(100)
sim1 = rtnew(1000,length(y1)-1, mean(y1), scale = sd(y1)/2)
sim2 = rtnew(1000,length(y2)-1, mean(y2), scale = sd(y2)/2)
    
delta_mean = sim1-sim2

  

    
  
print(mean(delta_mean))
print(mu_interval(delta_mean))
hist(delta_mean, main = " Delta of means", xlab = "Delta of mean", col = "lightblue")
```

## (c)

Write your answers here!

The probability for a single point in continous distributions is always
zero. The means are continously distributed, and this is the reason that
the probability is zero for the means of hardness being same across two
production lines.


